---
title: "Assignment"
author: "Sam Zinbarg"
output: html_document
date: "2/19/2023"
---

```{r library}
library(tidyverse)
library(ggplot2)
```
## Load and check data

### Problem 2
``` {r read in data}
gapminder <- read_delim("gapminder.csv")
gapminder

nrow(gapminder)
ncol(gapminder)
```
There are 13055 rows of data and 25 colummns

### Problem 3
``` {r sample data}
 gapminder %>% 
  sample_n(5)
```
The data appears to mostly be fine however there is some data that has missing values as there is some columns that have some NAs for some of the values

## Descriptive statistics

### Problem 1
``` {r counting countries}
n_distinct(gapminder$name, na.rm = TRUE)
n_distinct(gapminder$iso3, na.rm = TRUE)
n_distinct(gapminder$iso2, na.rm = TRUE)
```
### Problem 2
```{r names to iso2 and to iso3}
gapminder %>% 
  group_by(iso2) %>%
  filter(is.na(iso2)) %>% 
  distinct(name)

gapminder %>% 
  group_by(iso3) %>% 
  distinct(name)
```
### Problem 3
```{r min and max years}
gapminder %>% 
  summarize(min(time, na.rm = TRUE),
            max(time, na.rm = TRUE))
```

## CO2 Emissions

### Problem 1
```{r missing CO2 values}
gapminder %>% 
  group_by(time) %>% 
  summarize(na_co2 = sum(is.na(co2)), na_co2_PC = sum(is.na(co2_PC))) %>% 
  arrange(desc(na_co2), na_co2_PC)
```
The years with the most missing values is 2017, 2018, 2019.

### Problem 2
```{r plot of C02 overtime in U.S, China, India}

US <- gapminder %>% filter(name == "United States of America")
China <- gapminder %>% filter(name == "China")
India <- gapminder %>% filter(name == "India")
Australia <- gapminder %>%  filter(name == "Australia")
Kenya <- gapminder %>%  filter(name == "Kenya")


ggplot(data = gapminder) + 
  geom_line(data = US, aes(x = time, y = co2, color = "USA")) +
  geom_line(data = China, aes(x = time, y = co2, color = "China")) +
  geom_line(data = India, aes(x = time, y = co2, color = "India")) +
  geom_line(data = Australia, aes(x = time, y = co2, color = "Australia")) +
  geom_line(data = Kenya, aes(x = time, y = co2, color = "Kenya")) 
```

### Problem 3
``` {r same graph but with co2_PC}

ggplot(data = gapminder) + 
  geom_line(data = US, aes(x = time, y = co2_PC, color = "USA")) +
  geom_line(data = China, aes(x = time, y = co2_PC, color = "China")) +
  geom_line(data = India, aes(x = time, y = co2_PC, color = "India")) +
  geom_line(data = Australia, aes(x = time, y = co2_PC, color = "Australia")) +
  geom_line(data = Kenya, aes(x = time, y = co2_PC, color = "Kenya")) 
```

### Problem 4
```{r co2_PC for each region}
gapminder %>% 
  filter(!is.na(region)) %>% 
  group_by(region) %>% 
  summarize(ave_co2_PC = mean(co2_PC, na.rm = TRUE))
```

### Problem 5
``` {r barplot for ave co2_PC by region}
gapminder %>% 
  filter(!is.na(region), time == "1960" | time == "2016") %>% 
  group_by(region, time) %>% 
  summarize(ave_co2_PC = mean(co2_PC, na.rm = TRUE)) %>% 
  ggplot(aes(region, ave_co2_PC, fill = factor(time))) +
  geom_bar(stat = "identity", position = position_dodge())
```

### Problem 6
```{r comparing countries ave_co2 in 2019 by continent}
gapminder %>% 
  filter(!is.na(region), time == "2016", !is.na(co2_PC)) %>% 
  group_by(region) %>% 
  reframe(region, name, co2_PC) %>% 
  filter(rank(co2_PC) <= 3 | rank(desc(co2_PC)) <= 3) %>%
  arrange(co2_PC)
  
```


## GDP per captia

###Problem 1
``` {r scatterplot}
GDP_PC_lifeExpectancy_1960 <- gapminder %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), time =="1960", !is.na(region)) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region))
GDP_PC_lifeExpectancy_1960
```

### Problem 2
``` {r same scatterplot using data from 2019}
GDP_PC_lifeExpectancy_2019 <- gapminder %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), time =="2019", !is.na(region)) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region))
GDP_PC_lifeExpectancy_2019
```

### Problem 3


### Problem 4
``` {r average life expectancy by region}
gapminder %>% 
  filter(!is.na(region), !is.na(lifeExpectancy)) %>%
  filter(time == "1960" | time == "2019") %>% 
  group_by(region, time) %>%
  summarize(ave_lifeExpectancy = mean(lifeExpectancy))
```

### Problem 5
``` {r comparing LE growth from 1960 to 2019}
gapminder %>% 
  filter(!is.na(region), !is.na(lifeExpectancy)) %>%
  filter(time == "1960" | time == "2019") %>% 
  group_by(region, time) %>%
  summarize(ave_lifeExpectancy = mean(lifeExpectancy)) %>% 
  mutate(growth = ave_lifeExpectancy-lag(ave_lifeExpectancy)) %>% 
  arrange(-growth)
```

### Problem 6
``` {r histograms}
gapminder %>% 
  filter(!is.na(GDP_PC), time == "1960" | time == "2019") %>% 
  ggplot(aes(x = GDP_PC, fill = factor(time)))+
  geom_histogram(position = position_dodge())
```
### Problem 7
```{r ranking us in LE both 1960 and 2019}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name)) %>%
  filter(time == "1960" | time == "2019") %>%
  group_by(time) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  reframe(name, time, rank)
```
### Problem 8






